ExEl: Building an Elaborator Using Extensible Constraints

### 2024-02-28 paper/main.tex
switch to tex build by default
add a comment explaining what the changes are to the tex file
grammar
don't forget to actually add the header

\PassOptionsToPackage{unicode}{hyperref}

@@

\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}

@@

\documentclass[
  sigconf,
  screen,
  review]{acmart}

@@

\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother

@@

\widowpenalty10000
\clubpenalty10000

@@

In this work we focus on a part of the type-checker called the elaborator that transforms user-friendly surface syntax to a small well-behaved core theory.
Our core idea is to use an open datatype for constraints and plugin system for solvers.

@@

\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}

@@

  colorlinks=true,
  linkcolor={Maroon},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},

@@

\hypertarget{sec:introduction}{%
\section{Introduction}\label{sec:introduction}}

Strongly-typed languages allow us to catch certain classes of bugs at
compile-time by checking the implementation against the type-signature.
When the types are provided by the user they can be viewed as a form of
specification, constraining the behaviour of the programs. This comes
with the benefit of more static guarantees but with an increased toll on
the user to supply more precise information about the program. Many
languages choose to infer types, but another option is to use the
information in the types to infer parts of the program. This idea was
aptly worded by Conor McBride as ``Write more types and fewer
programs.''
\citep[chap.~2.1]{ptoopTypeInferenceThought2022, mcbrideEpigramPracticalProgramming2005}
Some examples of this include overloaded functions in Java, implicits in
Scala, and type classes in Haskell.

In dependently typed languages like Agda
\citep{norellPracticalProgrammingLanguage2007, theagdateamAgdaUserManual2023a},
Coq \citep{thecoqdevelopmentteamCoqProofAssistant2022} or Idris
\citep{bradyIdrisGeneralpurposeDependently2013} the types can be much
more precise. This gives us an opportunity to infer even larger parts of
the program from the type in the process of elaboration. Examples
include implicit arguments in Agda, implicit coercions in Coq, and
tactic arguments in Idris. The inference or ``solving'' can be not only
automatic but also interactive or partially automatic. Holes in Agda and
proof obligations in Coq are examples of interactive, while Canonical
Structures \citep{mahboubiCanonicalStructuresWorking2013} in Coq and
program-synthesis for holes in Haskell
\citep{koppelSearchingEntangledProgram2022} are partially automatic. All
these mechanisms use different algorithms and have various degrees of
extensibility. These are typically not very well isolated from each
other and can therefore interact in unexpected ways, for example in case
between implicits and instance search
\citep{agdausersPerformanceRegressionIssue2018}. This puts a toll on the
language developer to modulate the interactions and for the user to
understand these features.

In all of these examples, solvers evolved organically over time together
with the language. Coq historically struggled with similar issues in the
elaborator: for example, Canonical Structures was not properly
documented for 15 years \citep{mahboubiCanonicalStructuresWorking2013}.
Agda experimented with features baked into the core of the type system
-- like sized types -- which brought their own solver infrastructure
\citep{abelExtensionMartinLofType2016}. Lean 4 is aiming to allow the
users to develop new surface-level features
\citep{leonardodemouraLeanMetaprogramming2021} using elaboration monads
\citep{mouraLeanTheoremProver2021}, somewhat akin to elaborator
reflection in Idris
\citep{christiansenElaboratorReflectionExtending2016}, but Lean 3 was
built in a more conventional way \citep{demouraLeanTheoremProver2015a}.

One common piece of infrastructure needed by these inference algorithms
are metavariables, also known as ``existential variables''
\citep[chap.~2.2.1]{thecoqdevelopmentteamCoqProofAssistant2022}, which
represent as-of-yet unknown parts of the program. Another one is the
need is to constrain two terms to be equal, known as unification.
Metavariables and unification are heavily used throughout the compiler,
for inference of implicit arguments and for general type-checking,
making the compiler sensitive towards changes in unification algorithms.
Because of the complexity unification breaking changes are often
discovered only when ran against a large existing project on CI, like
\texttt{cubical} or \texttt{stdlib} for Agda or \texttt{unimath} for
Coq.

In this paper we propose a new architecture for an extensible elaborator
for dependently typed languages. The idea is to provide an API for
developers to tap into the elaboration procedure with custom solvers
that can manipulate metavariables and constraints placed on them. This
design separates the `what' the solvers are doing from the `when',
making the interaction points between different parts of the
type-checker explicit. As a result, this allows the developer to reason
more easily about exceptions and asynchronicity in the type-checker and
add new features in a more contained fashion. Practically, this means
that each feature is contained within one module, as opposed to being
spread around the codebase.

Contributions:

\begin{itemize}
\tightlist
\item
  We propose a new design blueprint for implementing a language that is
  extensible with new constraints and new solvers. It supports type
  classes (Section \ref{sec:case-typeclasses}), implicit arguments
  (Section \ref{sec:case-implicits}), implicit coercions and tactic
  arguments (Section \ref{sec:coercion-tactics}).
\item
  We propose a view on metavariables as communication channels for the
  solvers, drawing parallels with asynchronous programming primitives
  (Section \ref{sec:solvers-implementation}).
\item
  We decompose the usual components of a type-checker, like the unifier
  in Agda, into a suite of solvers which can be extended and interleaved
  by user-provided plugins (Section
  \ref{sec:constraints_and_unification}).
\item
  Following the blueprint, we present a prototype implementation of a
  dependently typed language available at
  \href{https://github.com/liesnikov/extensible-elaborator}{github.com/liesnikov/extensible-elaborator}.
\end{itemize}

\hypertarget{sec:unification_constraint_based_elaboration_and_design_challanges}{%
\section{Unification, constraint-based elaboration and design
challenges}\label{sec:unification_constraint_based_elaboration_and_design_challanges}}

Constraints have been an integral part of compilers for strongly-typed
languages for a while \citep{oderskyTypeInferenceConstrained1999}. For
example, both Haskell \citep{vytiniotisOutsideInModularType2011} and
Agda \citep[ chap.~3]{norellPracticalProgrammingLanguage2007} use
constraints extensively. In the former case, they are even reflected and
can be manipulated by the user
\citep[chap.~6.10.3]{orchardHaskellTypeConstraints2010a, ghcdevelopmentteamGHCUserGuide2022}.
This has proved to be a good design decision for GHC, as is reflected,
for example in a talk by \citet{peytonjonesTypeInferenceConstraint2019},
as well as in a few published sources
\citep{vytiniotisOutsideInModularType2011, peytonjonesPracticalTypeInference2007}.

In the land of dependently typed languages constraints are much less
principled. Agda has a family of constraints\footnote{\href{https://github.com/agda/agda/blob/v2.6.4/src/full/Agda/TypeChecking/Monad/Base.hs\#L1157-L1191}{./src/full/Agda/TypeChecking/Monad/Base.hs\#L1157-L1191}.
  Here and hencefoth we shorten the links in footnotes to paths in the
  repository, the source code can be found at
  \href{https://github.com/agda/agda/blob/v2.6.4/}{github.com/agda/agda/blob/v2.6.4/}.}
that grew organically, currently counting 19 constructors. Idris
technically has constraints\footnote{\href{https://github.com/idris-lang/Idris2/blob/e673d05a67b82591131e35ccd50fc234fb9aed85/src/Core/UnifyState.idr}{./src/Core/UnifyState.idr}
  at
  \href{https://github.com/idris-lang/Idris2/blob/e673d05a67b82591131e35ccd50fc234fb9aed85}{github.com/idris-lang/Idris2/blob/e673d0}},
with the only two constructors being equality constraints for two terms
and for two sequences of terms. The same\footnote{\href{https://github.com/leanprover/lean4/blob/0a031fc9bbb43c274bb400f121b13711e803f56c/src/Lean/Meta/Match/Basic.lean\#L161}{./src/Lean/Meta/Match/Basic.lean\#L161}
  at
  \href{https://github.com/leanprover/lean4/blob/0a031fc9bbb43c274bb400f121b13711e803f56c/}{github.com/leanprover/lean4/blob/0a031f}}
holds for Lean. These languages either use constraints in a restricted,
single use-case manner -- namely, for unification -- or in an ad-hoc
manner.

In this section we will demonstrate why a more methodical approach to
constraints will result in more robust elaborators across the board. We
go over three typical challenges that come up when building a compiler
for a dependently typed language and the way they are usually solved. We
cover unification of the base language and the complexity of managing
the state of the unifier in Section \ref{sec:conversion_checking}. Then
we take a look at different kinds of implicit arguments and their
implementation in Section \ref{sec:implicit-arguments}. We briefly touch
on the problem of extending the unifier in Section
\ref{sec:extending-unification}. Finally, we summarize the issues in
Section \ref{sec:summary-of-the-issues}.

\hypertarget{sec:conversion_checking}{%
\subsection{Unification in presence of
meta-variables}\label{sec:conversion_checking}}

As mentioned in the introduction, in the process of type-checking we use
unification to compare terms, which is notoriously hard to implement.
The complexity stems from the desire of compiler writers to implement
the most powerful unifier, while being limited by the fact that
higher-order unification is undecidable in general. Some of this
complexity is unavoidable, but we can manage it better by splitting up
the unifier into smaller modular parts. In practice, this means that one
does not have to fit together an always-growing unifier but can instead
write different cases separately.

An example from Agda's elaborator is type-driven unification, which is
spread between about a hundred functions and 2200 lines of
code\footnote{\href{https://github.com/agda/agda/blob/v2.6.4/src/full/Agda/TypeChecking/Conversion.hs}{./src/full/Agda/TypeChecking/Conversion.hs}}.
Each of the functions implements part of the ``business logic'' of the
unifier. But all of them contain a lot of code that deals with
bookkeeping related to metavariables and constraints:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  They have to throw and catch exceptions, driving the control flow of
  the unification.
\item
  They have to compute blockers that determine when a postponed
  constraint is retried.
\item
  They have to deal with cases where either or both of the sides
  equation or its type are either metavariables or terms whose
  evaluation is blocked on some metavariables.
\end{enumerate}

Concretely, this code uses functions like \texttt{noConstraints} and
\texttt{dontAssignMetas} which rely on specific behaviour of the
constraint solver system. Other functions like \texttt{abortIfBlocked},
\texttt{reduce} and \texttt{catchConstraint}/\texttt{patternViolation}
force the programmer to make a choice between letting the constraint
system handle blockers or doing it manually. These things are known to
be brittle and pose an increased mental overhead when making changes.

The unifier is heavily used throughout the type-checker: either as
function calls to \texttt{leqType} when type-checking terms and
\texttt{compareType} when type-checking applications, or as raised
constraints \texttt{ValueCmp} and \texttt{SortCmp} from
\texttt{equalTerm} while checking applications or definitions,
\texttt{ValueCmpOnFace} from \texttt{equalTermOnFace} again while
checking applications. At the same it is unintuitive and full of
intricacies as indicated by multiple comments\footnote{\href{https://github.com/agda/agda/blob/v2.6.4/src/full/Agda/TypeChecking/Conversion.hs\#L484-L485}{./src/full/Agda/TypeChecking/Conversion.hs\#\#L484-L485},
  \href{https://github.com/agda/agda/blob/v2.6.4/src/full/Agda/TypeChecking/Conversion.hs\#L541-L549}{./src/full/Agda/TypeChecking/Conversion.hs\#L541-L549}}.

We would like the compiler-writer to separate the concerns of managing
constraints and blockers from the actual logic of the comparison
function. In fact, if we zoom in on the \texttt{compareAtom} function,
the core can be expressed in about 20 lines\footnote{\href{https://github.com/agda/agda/blob/v2.6.4/src/full/Agda/TypeChecking/Conversion.hs\#L550-L599}{./src/full/Agda/TypeChecking/Conversion.hs\#L550-L599}}
of simplified code, stripping out size checks, cumulativity, polarity,
and forcing. This is precisely what we would like the developer to
write.

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{case}\NormalTok{ (m, n) }\KeywordTok{of}
\NormalTok{  (}\DataTypeTok{Lit}\NormalTok{ l1, }\DataTypeTok{Lit}\NormalTok{ l2) }\OperatorTok{|}\NormalTok{ l1 }\OperatorTok{==}\NormalTok{ l2 }\OtherTok{{-}\textgreater{}} \FunctionTok{return}\NormalTok{ ()}
\NormalTok{  (}\DataTypeTok{Var}\NormalTok{ i es, }\DataTypeTok{Var}\NormalTok{ i\textquotesingle{} es\textquotesingle{}) }\OperatorTok{|}\NormalTok{ i }\OperatorTok{==}\NormalTok{ i\textquotesingle{} }\OtherTok{{-}\textgreater{}} \KeywordTok{do}
\NormalTok{      a }\OtherTok{\textless{}{-}}\NormalTok{ typeOfBV i}
\NormalTok{      compareElims [] [] a (var i) es es\textquotesingle{}}
\NormalTok{  (}\DataTypeTok{Con}\NormalTok{ x ci xArgs, }\DataTypeTok{Con}\NormalTok{ y \_ yArgs) }\OperatorTok{|}\NormalTok{ x }\OperatorTok{==}\NormalTok{ y }\OtherTok{{-}\textgreater{}} \KeywordTok{do}
\NormalTok{      t\textquotesingle{} }\OtherTok{\textless{}{-}}\NormalTok{ conType x t}
\NormalTok{      compareElims t\textquotesingle{} (}\DataTypeTok{Con}\NormalTok{ x ci []) xArgs yArgs}
  \OperatorTok{...}
\end{Highlighting}
\end{Shaded}

The functions described above are specific to Agda but in other major
languages we can find similar problems with unifiers being large modules
that are hard to understand. The sizes of modules with unifiers are as
follows: Idris (1.5kloc\footnote{\href{https://github.com/idris-lang/Idris2/blob/102d7ebc18a9e881021ed4b05186cccda5274cbe/src/Core/Unify.idr}{./src/Core/Unify.idr}}),
Lean (1.8kloc\footnote{\href{https://github.com/leanprover/lean4/blob/75252d2b85df8cb9231020a556a70f6d736e7ee5/src/Lean/Meta/ExprDefEq.lean}{./src/Lean/Meta/ExprDefEq.lean}}),
Coq (1.8kloc\footnote{\href{https://github.com/coq/coq/blob/b35c06c3ab3ed4911311b4a9428a749658d3eff1/pretyping/evarconv.mli}{./pretyping/evarconv.mli}
  at
  \href{github.com/coq/coq/blob/b35c06c3ab3ed4911311b4a9428a749658d3eff1/}{github.com/coq/coq/blob/b35c06}}).
For Haskell, which is not a dependently typed language yet, but does
have a constraints system
\citep{peytonjonesTypeInferenceConstraint2019}, this number is at
2kloc\footnote{\href{https://gitlab.haskell.org/ghc/ghc/-/blob/b81cd709df8054b8b98ac05d3b9affcee9a8b840/compiler/GHC/Core/Unify.hs}{./compiler/GHC/Core/Unify.hs}
  at
  \href{https://gitlab.haskell.org/ghc/ghc/-/blob/b81cd709df8054b8b98ac05d3b9affcee9a8b840}{gitlab.haskell.org/ghc/ghc/-/blob/b81cd709d}}.

\hypertarget{sec:implicit-arguments}{%
\subsection{Type-checking function application in presence of implicit
arguments}\label{sec:implicit-arguments}}

During the type-checking of function application there may be different
kinds of implicit arguments to infer, for example, instance arguments,
or tactic arguments. If we start from a simple case of type-checking an
application of a function symbol to regular arguments, every next
extension needs to be handled as a special case.

Take Agda -- when checking an application of implicit
arguments\footnote{\href{https://github.com/agda/agda/blob/v2.6.4/src/full/Agda/TypeChecking/Implicit.hs\#L96-L128}{./src/full/Agda/TypeChecking/Implicit.hs\#L96-L128}}
we already have to carry the information on how the argument will be
resolved and then create a specific kind of metavariable\footnote{\href{https://github.com/agda/agda/blob/v2.6.4/src/full/Agda/TypeChecking/Implicit.hs\#L130-L149}{./src/full/Agda/TypeChecking/Implicit.hs\#L130-L149}}
\citep[chap.~3]{norellPracticalProgrammingLanguage2007} for each of
those cases.

For handling \texttt{auto} variables, Idris 2
\citep[chap.~13.1]{theidristeamIdrisTutorial2021} has to essentially
inline the search procedure through a chain of elaboration function
calls (\texttt{checkApp} to \texttt{checkAppWith} to
\texttt{checkAppWith\textquotesingle{}}) to
\texttt{makeAutoImplicit}\footnote{\href{https://github.com/idris-lang/Idris2/blob/870bc824371d504a03af937f326216302210a875/src/TTImp/Elab/App.idr\#L224-L241}{./src/TTImp/Elab/App.idr\#L224-L241}}.
This can accommodate interfaces (or type classes), but one can imagine
that if a different kind of implicit was added, like tactic arguments or
Canonical Structures, we would have to inline the search again,
requiring a non-trivial modification to the elaboration mechanism.

While the codebases above show that it is certainly possible to extend
languages with new features if the language was not written with
extensibility in mind, this can lead to rather ad hoc solutions.

\hypertarget{sec:extending-unification}{%
\subsection{Extending unification}\label{sec:extending-unification}}

While writing a unifier is hard enough as it is, at times the developers
might want to give their users ability to extend the unification
procedure.

Canonical Structures
\citep{saibiOutilsGeneriquesModelisation1999, mahboubiCanonicalStructuresWorking2013}
was already mentioned as it is in the overlap between type classes and
unification hints. Adding it to a language that does not support them
requires an extension of the unification algorithm with a rule that says
that projection from a canonical structure is an injective function
\citep[eq. 1]{mahboubiCanonicalStructuresWorking2013}.

One could also provide means to do so manually in a more general case,
by allowing users to declare certain symbols as injective. This is one
of the features requested by the Agda users
\citep{agdausersInjectiveUnificationPragma2023}.

Another example of this can be adding rules of associativity and
commutativity to the unifier, as described in the thesis by
\citet{holtenDependentTypeCheckingModulo2023}. It required 2000 lines of
code\footnote{\href{https://github.com/LHolten/agda-commassoc/tree/defenitional-commutativity}{github.com/LHolten/agda-commassoc/tree/defenitional-commutativity}}.
We would like to make changes such as this more feasible.

\hypertarget{sec:summary-of-the-issues}{%
\subsection{Summary of the issues}\label{sec:summary-of-the-issues}}

The examples above show that when building a dependently typed language
the core might be perfectly elegant and simple, but the features that
appear on top of it complicate the design.

One can also observe that while the code in existing implementations
might rely on constraints, the design at large does not put them at the
centre of the picture and instead views them primarily as a gadget.
Agda's constraint solver\footnote{\href{https://github.com/agda/agda/blob/v2.6.4/src/full/Agda/TypeChecking/Constraints.hs\#L247-L298}{src/full/Agda/TypeChecking/Constraints.hs\#L247-L298}}
relies on the type-checker to call it at the point where it is needed
and has to be carefully engineered to work with the rest of the
codebase.

\hypertarget{sec:what-is-our-design-bringing-into-the-picture}{%
\section{A blueprint for extensible
elaborators}\label{sec:what-is-our-design-bringing-into-the-picture}}

Our idea for a new design is to shift focus more towards the constraints
themselves:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  We give a stable API for raising constraints that can be called by the
  type-checker, essentially creating an ``ask'' to be fulfilled by the
  solvers. This is not dissimilar to the idea of mapping object-language
  unification variables to host-language ones as done by
  \citet{guidiImplementingTypeTheory2017}, view of the ``asks'' as a
  general effect \citep[chap.~4.4]{bauerEqualityCheckingGeneral2020}, or
  communication between actors \citep{allaisTypOSOperatingSystem2022a}.
\item
  To make the language more modular, we make constraints an extensible
  data type in the style of \citet{swierstraDataTypesCarte2008} and give
  an API to define new solvers with the ability to specify what kinds of
  constraints they can solve.
\end{enumerate}

In the examples in this paper, we follow the bidirectional style of
type-checking, but in practice, the design decisions are agnostic of the
underlying system, as long as it adheres to the principle of stating the
requirements on terms in terms of raising a constraint and not by, say,
pattern-matching on a concrete term representation.

From a birds-eye view the architecture looks as depicted in Figure
\ref{architecture-figure}. The type-checking begins by initializing
state and doing the syntax traversal. The traversal raises the
constraints, and for the moment, the constraints are simply stored. As
soon as we finish the traversal of some block (one declaration in our
case), the solver dispatcher is called. It goes over the set of
constraints, and for each active constraint calls different plugins to
make an attempt at solving it. Each plugin, whether user-supplied
(\texttt{Plugin\ A}) or provided by us (\texttt{unification}) consist of
a handler and a solver. The handler determines if the plugin can
potentially solve a constraint, if so the dispatcher runs the
corresponding solver. All components have some read access to the state,
including handlers which might for example verify that there are no
extra constraints on the metavariable. For the write access: syntax
traversal writes new metavariables to the state and elaborated
definitions; the solver dispatcher writes updated meta-information;
solvers write solutions to the metavariables and can raise new
constraints.

\begin{figure*}
  \center

  \includegraphics[width=\textwidth]{architecture-diagram.png}

  \caption{Architecture diagram}
  \label{architecture-figure}
\end{figure*}

For the moment we need to recompile the project in order to include new
plugins, but this is not necessary and a system that dynamically loads
plugins is possible to implement in a way that is similar to GHC
Plugins\footnote{\href{https://mpickering.github.io/plugins.html}{mpickering.github.io/plugins.html}}
or Accelerate \footnote{\href{https://github.com/tmcdonell/accelerate-llvm/blob/master/accelerate-llvm-native/src/Data/Array/Accelerate/LLVM/Native/Link/Runtime.hs\#L40}{github.com/tmcdonell/accelerate-llvm/blob/master/accelerate-llvm-native/src/Data/Array/Accelerate/LLVM/Native/Link/Runtime.hs}}
\citep{mcdonellTypesafeRuntimeCode2015}.

\hypertarget{sec:bidirectional}{%
\section{Dependently typed calculus and bidirectional
typing}\label{sec:bidirectional}}

In this section, we describe the core of the type system we implement.
We take pi-forall \citep{weirichImplementingDependentTypes2022} as a
basis for the system and extend it with metavariables in core syntax and
implicit arguments in the surface syntax. However, for all other
purposes, we leave the core rules intact and therefore, the core
calculus too.

\hypertarget{basic-language-and-rules}{%
\subsection{Basic language and rules}\label{basic-language-and-rules}}

This is a dependently typed calculus that includes Pi, Sigma and indexed
inductive types.

Here is the internal syntax term data type, apart from service
constructor omissions, like a placeholder \texttt{TRUSTME}.

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{data} \DataTypeTok{Term} \OtherTok{=}
  \CommentTok{{-}{-} type of types Type}
    \DataTypeTok{Type}
  \CommentTok{{-}{-} variables x}
  \OperatorTok{|} \DataTypeTok{Var} \DataTypeTok{TName}
  \CommentTok{{-}{-} abstraction \textbackslash{}x. a}
  \OperatorTok{|} \DataTypeTok{Lam}\NormalTok{ (}\DataTypeTok{Bind} \DataTypeTok{TName} \DataTypeTok{Term}\NormalTok{)}
  \CommentTok{{-}{-} application a b}
  \OperatorTok{|} \DataTypeTok{App} \DataTypeTok{Term} \DataTypeTok{Arg}
  \CommentTok{{-}{-} function type (x : A) {-}\textgreater{} B}
  \OperatorTok{|} \DataTypeTok{Pi} \DataTypeTok{Epsilon} \DataTypeTok{Type}\NormalTok{ (}\DataTypeTok{Bind} \DataTypeTok{TName} \DataTypeTok{Type}\NormalTok{)}
  \CommentTok{{-}{-} Sigma{-}type \{ x : A | B \}}
  \OperatorTok{|} \DataTypeTok{Sigma} \DataTypeTok{Term}\NormalTok{ (}\DataTypeTok{Bind} \DataTypeTok{TName} \DataTypeTok{Term}\NormalTok{)}
  \OperatorTok{|} \DataTypeTok{Prod} \DataTypeTok{Term} \DataTypeTok{Term}
  \OperatorTok{|} \DataTypeTok{LetPair} \DataTypeTok{Term}\NormalTok{ (}\DataTypeTok{Bind}\NormalTok{ (}\DataTypeTok{TName}\NormalTok{, }\DataTypeTok{TName}\NormalTok{) }\DataTypeTok{Term}\NormalTok{)}
  \CommentTok{{-}{-} Equality type  a = b}
  \OperatorTok{|} \DataTypeTok{TyEq} \DataTypeTok{Term} \DataTypeTok{Term}
  \OperatorTok{|} \DataTypeTok{Refl}
  \OperatorTok{|} \DataTypeTok{Subst} \DataTypeTok{Term} \DataTypeTok{Term}
  \OperatorTok{|} \DataTypeTok{Contra} \DataTypeTok{Term}
  \CommentTok{{-}{-} inductive datatypes}
  \OperatorTok{|} \DataTypeTok{TCon} \DataTypeTok{TCName}\NormalTok{ [}\DataTypeTok{Arg}\NormalTok{] }\CommentTok{{-}{-} types (fully applied)}
  \OperatorTok{|} \DataTypeTok{DCon} \DataTypeTok{DCName}\NormalTok{ [}\DataTypeTok{Arg}\NormalTok{] }\CommentTok{{-}{-} terms (fully applied)}
  \OperatorTok{|} \DataTypeTok{Case} \DataTypeTok{Term}\NormalTok{ [}\DataTypeTok{Match}\NormalTok{]}
    \CommentTok{{-}{-} metavariables}
  \OperatorTok{|} \DataTypeTok{MetaVar} \DataTypeTok{MetaClosure}
\end{Highlighting}
\end{Shaded}

Equality is not defined as a regular inductive type, but is instead
built-in. The user has access to the type and term constructor, but not
the ability to pattern-matching on it. Instead the language provides a
\texttt{subst} primitive of type
\texttt{(A\ x)\ -\textgreater{}\ (x=y)\ -\textgreater{}\ A\ y} and
\texttt{contra} that takes an equality of two different inductive type
constructors and produces an element of any type.

On top of the above the language includes indexed inductive datatypes
and case-constructs for their elimination. Indexed inductive datatypes
are encoded as parameterised datatypes with an equality argument
constraining the index.

As for metavariables \texttt{MetaVar}: as mentioned in the introduction,
they are placeholders in the syntax tree (AST) that are produced in the
process of elaboration. Metavariables do not appear in the surface
syntax as they are not created by the user. In this paper we implement
metavariables in the contextual style, as described by
\citet{abelHigherOrderDynamicPattern2011}, therefore they have to carry
around a closure of type \texttt{MetaClosure}.

\hypertarget{syntax-traversal}{%
\subsection{Syntax traversal}\label{syntax-traversal}}

We implement the core of the elaborator as a bidirectional syntax
traversal, raising a constraint every time we need to assert something
about the type.

This includes the expected use of unification constraints, like in case
we enter checking mode while the term should be inferred:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{checkType tm ty }\OtherTok{=} \KeywordTok{do}
\NormalTok{  (etm, ty\textquotesingle{}) }\OtherTok{\textless{}{-}}\NormalTok{ inferType tm}
\NormalTok{  constrainEquality ty\textquotesingle{} ty }\DataTypeTok{I.Type}
  \FunctionTok{return}\NormalTok{ etm}
\end{Highlighting}
\end{Shaded}

Any time we want to decompose the type provided in checking mode we pose
a constraint that guarantees the type being in a specific shape:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{checkType (}\DataTypeTok{S.Lam}\NormalTok{ lam) ty }\OtherTok{=} \KeywordTok{do}
\NormalTok{  mtyA }\OtherTok{\textless{}{-}}\NormalTok{ createMetaTerm}
\NormalTok{  mtx }\OtherTok{\textless{}{-}}\NormalTok{ createUnknownVar}
\NormalTok{  mtyB }\OtherTok{\textless{}{-}}\NormalTok{ extendCtx (}\DataTypeTok{I.TypeSig}\NormalTok{ (}\DataTypeTok{I.Sig}\NormalTok{ mtx mtyA))}
\NormalTok{                    (createMetaTerm)}
  \KeywordTok{let}\NormalTok{ mbnd }\OtherTok{=}\NormalTok{ bind mtx mtyB}
  \KeywordTok{let}\NormalTok{ metaPi }\OtherTok{=} \DataTypeTok{I.Pi}\NormalTok{ mtyA mbnd}

\NormalTok{  constrainEquality ty metaPi }\DataTypeTok{I.Type}
  \CommentTok{{-}{-} rest of the traversal can now use mtyA and mbnd}
  \OperatorTok{...}
\end{Highlighting}
\end{Shaded}

At certain points we have to raise a constraint which has an associated
continuation. Like for checking the type of a data constructor -- the
part of the program that comes as an argument to
\texttt{constrainTConAndFreeze} will be suspended (or ``blocked'') until
the meta is solved with something of shape \texttt{TCon\ \_\ \_}.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{checkType t}\OperatorTok{@}\NormalTok{(}\DataTypeTok{S.DCon}\NormalTok{ c args) ty }\OtherTok{=} \KeywordTok{do}
\NormalTok{  elabpromise }\OtherTok{\textless{}{-}}\NormalTok{ createMetaTerm}
\NormalTok{  constrainTConAndFreeze ty }\OperatorTok{$} \KeywordTok{do}
\NormalTok{    mty }\OtherTok{\textless{}{-}}\NormalTok{ substMetas ty}
    \KeywordTok{case}\NormalTok{ mty }\KeywordTok{of}
\NormalTok{      (}\DataTypeTok{I.TCon}\NormalTok{ tname params) }\OtherTok{{-}\textgreater{}} \KeywordTok{do}
        \OperatorTok{...}
\NormalTok{      \_ }\OtherTok{{-}\textgreater{}} \FunctionTok{error} \StringTok{"impossible"}
\end{Highlighting}
\end{Shaded}

We add one more case to the elaborator for implicit arguments.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{checkType (}\DataTypeTok{Implicit}\NormalTok{) ty }\OtherTok{=} \KeywordTok{do}
\NormalTok{  m }\OtherTok{\textless{}{-}}\NormalTok{ createMetaTerm}
\NormalTok{  raiseConstraint }\OperatorTok{$} \DataTypeTok{FillInTheMeta}\NormalTok{ m ty}
  \FunctionTok{return}\NormalTok{ m}
\end{Highlighting}
\end{Shaded}

This is used for implicit arguments as will be described in more detail
in Section \ref{sec:case-implicits}. The syntax traversal does not need
to know anything at all about the implicits. The only thing we require
is that the elaboration of the argument is called with the type
information available. This corresponds to how in bidirectional typing
function application is done in the inference mode but the arguments are
processed in checking mode.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{inferType (}\DataTypeTok{App}\NormalTok{ t1 t2) }\OtherTok{=} \KeywordTok{do}
\NormalTok{  (et1, }\DataTypeTok{Pi}\NormalTok{ tyA tyB) }\OtherTok{\textless{}{-}}\NormalTok{ inferType t1}
\NormalTok{  et2 }\OtherTok{\textless{}{-}}\NormalTok{ checkType t2 tyA}
  \FunctionTok{return}\NormalTok{ (}\DataTypeTok{App}\NormalTok{ et1 et2, subst tyB et2)}
\end{Highlighting}
\end{Shaded}

\hypertarget{sec:constraints_and_unification}{%
\section{Constraints and
unification}\label{sec:constraints_and_unification}}

In Section \ref{sec:bidirectional} we described the syntax traversal
part of the elaborator, which generates the constraints. In this section
we will go over the constraint datatype needed for the base language,
how unification is implemented and how we can extend the unification
procedure.

\hypertarget{base-constraints}{%
\subsection{Base constraints}\label{base-constraints}}

The datatype of constraints is open which means the user can write a
plugin to extend it. However, we offer a few constraints out of the box
to be able to type-check the base language.

For the purposes of the base language it suffices to have the following.

\begin{itemize}
\item
  A constraint that enforces equality of two terms of a given type

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{{-}{-} two terms given should be equal}
\KeywordTok{data} \DataTypeTok{EqualityConstraint}\NormalTok{ e }\OtherTok{=}
     \DataTypeTok{EqualityConstraint} \DataTypeTok{Syntax.Term} \DataTypeTok{Syntax.Term}
                        \DataTypeTok{Syntax.Type}
                        \DataTypeTok{Sytax.MetaVarId}
\end{Highlighting}
\end{Shaded}
\item
  A constraint that ensures that a metavariable is resolved eventually:

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{{-}{-} this terms has to be filled in}
\KeywordTok{data} \DataTypeTok{FillInTheTerm}\NormalTok{ e }\OtherTok{=}
     \DataTypeTok{FillInTheTerm} \DataTypeTok{Syntax.Term}
\NormalTok{                   (}\DataTypeTok{Maybe} \DataTypeTok{Syntax.Type}\NormalTok{)}
\end{Highlighting}
\end{Shaded}
\item
  Lastly, a constraint which ensures that a term is a type constructor:

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{{-}{-} the term passed to the constraint}
\CommentTok{{-}{-} should be a type constructor}
\KeywordTok{data} \DataTypeTok{TypeConstructorConstraint}\NormalTok{ e }\OtherTok{=}
     \DataTypeTok{TypeConstructorConstraint} \DataTypeTok{Syntax.Type}
\end{Highlighting}
\end{Shaded}
\end{itemize}

The type-checker raises them supplying the information necessary, but
agnostic of how they will be solved.

\hypertarget{sec:solvers-interface}{%
\subsection{Interface of the solvers}\label{sec:solvers-interface}}

On the solver side we provide a suite\footnote{In the prototype we
  implement a subset of all unification rules, here are they listed:
  identityPlugin, propagateMetasEqPlugin, reduceLeftPlugin,
  reduceRightPlugin, leftMetaPlugin, rightMetaPlugin,
  typeConstructorPlugin, typeConstructorWithMetasPlugin,
  piEqInjectivityPlugin, tyEqInjectivityPlugin, consInjectivityPlugin,
  typeInjectivityPlugin, unificationStartMarker, unificationEndMarker}
of solvers for unification that handle different cases of the problem.

>>>>

@@

@@

\documentclass[sigconf, screen]{acmart}

@@

@@

\widowpenalty=10000
\clubpenalty=10000
\displaywidowpenalty=10000

@@

In this work we focus on a part of the type-checker called the elaborator that transforms user-friendly surface syntax into a small well-behaved core theory.
Our core idea is to use an open datatype for constraints and a plugin system for solvers.

@@

%\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}

@@

  hidelinks,

@@

\input{body.tex}
